---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CausalState

<!-- badges: start -->
[![R-CMD-check](https://github.com/YOUR-GITHUB-HANDLE/CausalState/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/YOUR-GITHUB-HANDLE/CausalState/actions/workflows/R-CMD-check.yaml)
[![License: AGPL v3](https://img.shields.io/badge/License-AGPL_v3-blue.svg)](https://www.gnu.org/licenses/agpl-3.0)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**CausalState** provides Sequential Doubly Robust (SDR) and iterative Targeted Maximum Likelihood (iTMLE) estimators for longitudinal modified treatment policies in settings with absorbing state transitions — for example, ICU, ward, or emergency department care episodes where treatment is applicable while in the active state and becomes structurally inapplicable after discharge or death.

The package complements [`lmtp`](https://github.com/nt-williams/lmtp) by handling settings where the timing of state transitions is itself endogenous to the treatment policy.

## Scope

Designed for **medium-horizon** care episodes (5–10 daily blocks). Beyond that, cumulative density-ratio products tend to collapse positivity regardless of estimator choice.

## Features

- Sequential Doubly Robust (SDR) estimator with k-fold cross-fitting
- Iterative TMLE (iTMLE) estimator
- First-class handling of single absorbing transitions (alive-discharge vs death as competing terminal events)
- Asymmetric g- and Q-model regularisation defaults (sharper g, more-regularised Q)
- Pluggable SuperLearner ensembles via included custom learner wrappers (`SL.tgt.*`, `SL.tmle_*`)
- Cluster bootstrap standard errors

## Installation

You can install the development version of `CausalState` from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("YOUR-GITHUB-HANDLE/CausalState")
```

## Quick example

```{r example, eval = FALSE}
library(CausalState)

# Coming soon — see vignettes/getting-started.Rmd
```

## Citation

Please cite this package as:

```
Blank S (2026). CausalState: SDR and iTMLE for State-Aware Longitudinal
Modified Treatment Policies. R package version 0.0.0.9000.
https://github.com/YOUR-GITHUB-HANDLE/CausalState
```

## License

AGPL-3. See `LICENSE.md` for details.
