---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CausalState

<!-- badges: start -->
[![License: AGPL v3](https://img.shields.io/badge/License-AGPL_v3-blue.svg)](https://www.gnu.org/licenses/agpl-3.0)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**CausalState** provides Sequential Doubly Robust (SDR) and
infinite-dimensional Targeted Maximum Likelihood (iTMLE) estimators for
longitudinal modified treatment policies (MTPs) in care-episode settings
where patients can transition irreversibly out of an active state -- for
example, ICU discharge or in-hospital death.

The distinctive feature of the package is that the MTP can **shift the
transition dynamics themselves**, not only the terminal outcome.  A policy
that shortens ICU stay changes both when patients leave and what outcomes
they experience after leaving; CausalState handles both simultaneously.

## Problem setting

The estimators apply when:

- Data are in **long format**, one row per patient per time point, with
  rows present only while the patient is in the active state.
- At each time point a patient can remain, discharge (alive exit), or die
  (dead exit); no return to the active state is possible.
- The treatment policy of interest is a **modified treatment policy** (MTP):
  a data-adaptive shift of the observed treatment rather than a static
  intervention.
- Interest is in the **mean counterfactual outcome** under the MTP, estimated
  from observational or trial data.

The package is designed for **medium-horizon** episodes of 5--10 time
points.  Beyond roughly ten points, cumulative density-ratio products tend
to collapse positivity regardless of estimator choice.

## Estimators

Both estimators are sequentially doubly robust (2^K-robust, Luedtke et al.
2017): consistent whenever, at each time point, either the treatment model
or the outcome model is correctly specified.

| | SDR (`sdr()`) | iTMLE (`itmle()`) |
|---|---|---|
| Update step | EIF pseudo-outcome (Diaz et al. 2021) | Infinite-dimensional TMLE fluctuation (Luedtke et al. 2017) |
| SE formula | Centered: `sd(IC) / sqrt(n)`, E[IC] = 0 by construction | Second-moment: `sqrt(mean(IC^2) / n)`, conservative when targeting is near-convergence |
| Natural-course run | Collapses to `mean(Y)` -- not a model check | Does not collapse -- genuine Q-model calibration check |
| Extra inputs | None beyond Q/g libraries | Targeting SL library (`sl_tmle`) |

`qreg()` is a pure Q-recursion plug-in (no update, no DR guarantees) included
as a weight-independent sensitivity check.

## Workflow

```
density_ratio()  ->  sdr() / itmle() / qreg()
```

`density_ratio()` must run first.  It fits per-time-point treatment
classification models, computes the instantaneous density ratios
r_t = dP~(A_t|H_t) / dP(A_t|H_t), and packages them with fold assignments
that the downstream estimators inherit.  Running it once and passing the
result to multiple estimators is the intended pattern.

## Quick example

```{r example, eval = FALSE}
library(CausalState)
library(SuperLearner)

# --- 1. Define the policy -------------------------------------------------
# Soft upward shift: nudge treatment probability up by 0.3, capped at 1
policy_up <- function(D_block, t, a_names) {
  out <- D_block[, ..a_names, drop = FALSE]
  out[[a_names[1]]] <- pmin(D_block[[a_names[1]]] + 0.3, 1)
  out
}

# --- 2. Density ratios ----------------------------------------------------
sl_lib <- c("SL.mean", "SL.glm")   # replace with richer library in practice

wr <- density_ratio(
  df              = patient_data,   # long-format data frame
  a_names         = "A",
  tmax            = 7L,
  baseline        = c("age", "sex"),
  tv_names        = c("L1", "L2"),
  sl_g            = sl_lib,
  k               = 5L,
  inner_v         = 5L,
  v               = 5L,
  seed            = 1L,
  id              = "id",
  time            = "time",
  policy_spec_fun = policy_up
)

# --- 3a. SDR estimate -----------------------------------------------------
res_sdr <- sdr(
  df              = patient_data,
  weight_object   = wr,
  tmax            = 7L,
  id              = "id", time = "time",
  alive           = "alive", in_state = "in_state",
  y               = "Y",
  baseline        = c("age", "sex"),
  tv_names        = c("L1", "L2"),
  a_names         = "A",
  sl_remain       = sl_lib, sl_death = sl_lib,
  sl_recursive    = sl_lib, sl_y = sl_lib,
  outcome_family  = "binomial",
  k               = 5L, inner_v = 5L,
  seed            = 1L,
  policy_spec_fun = policy_up
)

cat(sprintf("SDR:   psi = %.3f  SE = %.3f  95%% CI [%.3f, %.3f]\n",
            res_sdr$psi, res_sdr$se, res_sdr$ci[1], res_sdr$ci[2]))

# --- 3b. iTMLE estimate ---------------------------------------------------
res_itmle <- itmle(
  df              = patient_data,
  weight_object   = wr,
  tmax            = 7L,
  id              = "id", time = "time",
  alive           = "alive", in_state = "in_state",
  y               = "Y",
  baseline        = c("age", "sex"),
  tv_names        = c("L1", "L2"),
  a_names         = "A",
  sl_remain       = sl_lib, sl_death = sl_lib,
  sl_recursive    = sl_lib, sl_y = sl_lib,
  sl_target       = sl_tmle,
  outcome_family  = "binomial",
  k               = 5L, inner_v = 5L,
  seed            = 1L,
  policy_spec_fun = policy_up
)

cat(sprintf("iTMLE: psi = %.3f  SE = %.3f  95%% CI [%.3f, %.3f]\n",
            res_itmle$psi, res_itmle$se, res_itmle$ci[1], res_itmle$ci[2]))

# --- 4. Risk difference ---------------------------------------------------
ctr <- contrast(res_sdr, res_sdr_nat)   # intervention vs natural course
ctr$RD; ctr$ci_RD
```

## Key design choices

**Asymmetric g/Q regularisation.**  Treatment models (g) are tuned sharper
for accurate density-ratio discrimination; outcome models (Q) are more
heavily regularised to suppress recursion noise.  This asymmetry is the
primary tuning lever.

**Uniform clipping.**  Every SL prediction across the entire pipeline (g and
Q models at every time point) is clipped to [bounds, 1-bounds] using the
same `bounds` parameter (default 1e-5).  The only exception is the
Wu-Benkeser direct density-ratio metalearner, which clips in density-ratio
space via `dr_floor` -- see `vignette("wb-metalearner")`.

**Weight reuse.**  `density_ratio()` is designed to be run once and shared
across `sdr()`, `itmle()`, and competing-event variants.  Trimming is applied
globally at consumption time, so all estimators that share a weight object
operate on identically trimmed weights and produce directly comparable
estimates.

**Parallelism via `mclapply`.**  Process-level parallelism is available at
the fold level (`parallel = TRUE`) and within-fold regression level
(`reg_workers`).  Does not work on Windows.  Set BLAS and learner thread
counts to 1 when enabling process-level parallelism to avoid
oversubscription.

## Installation

```r
# install.packages("remotes")
remotes::install_github("sebastiaan-blank/CausalState")
```

## Getting started

See `vignette("getting-started")` for a complete worked example with a
simulated ICU dataset, including data structure, policy definition,
SuperLearner library choices, diagnostics, and a competing-event analysis.

## Citation

```
Blank S (2026). CausalState: SDR and iTMLE for State-Aware Longitudinal
Modified Treatment Policies. R package version 0.9.0.
https://github.com/sebastiaan-blank/CausalState
```

## References

Diaz I, Williams N, Hoffman KL, Schenck EJ (2021). Nonparametric Causal
Effects Based on Longitudinal Modified Treatment Policies. *JASA*
118(542):846-857.

Luedtke AR, Sofrygin O, van der Laan MJ, Carone M (2017). Sequential Double
Robustness in Right-Censored Longitudinal Models. arXiv:1705.02459.

Williams NT, Diaz I (2023). lmtp: An R package for estimating the causal
effects of modified treatment policies. *Observational Studies*.

## License

AGPL-3. See `LICENSE.md` for details.
