---
title: "Wu-Benkeser density-ratio metalearner"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wu-Benkeser density-ratio metalearner}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

By default, `density_ratio()` estimates each instantaneous density ratio
r_t(h, a) = dP~(A_t = a | H_t = h) / dP(A_t = a | H_t = h) via a binary
classification approach: a SuperLearner is trained to distinguish natural-course
from shifted-course observations, and the ratio is recovered as
r = p-hat / (1 - p-hat) where p-hat is the predicted probability of belonging
to the shifted arm.

Setting `dr_sl = TRUE` activates the **Wu-Benkeser (WB) metalearner**
(`method.WB_dr()`), which combines base learners differently: instead of the
default non-negative least squares (NNLS) on the probability scale, it
minimises a log-density-ratio loss directly in density-ratio space.

---

## Standard approach: classify then convert

In the standard (`dr_sl = FALSE`) pathway:

1. Observations from the natural-course arm are labelled 0; observations from
   the shifted arm are labelled 1.  The combined dataset is passed to a binary
   SuperLearner.
2. Each base learner outputs a probability p-hat in (0, 1).
3. NNLS combines the per-learner predictions on the probability scale.
4. The density ratio is recovered as r = p-hat-ensemble / (1 - p-hat-ensemble).
5. Predictions are clipped to [bounds, 1 - bounds] before conversion (where
   `bounds` is the user-supplied clipping parameter, default 1e-5).

This is the approach used in the `lmtp` package and is the safe default.
Standard SuperLearner wrappers (`SL.glm`, `SL.xgboost`, `SL.earth`, etc.)
all work here because they return probabilities.

---

## WB approach: combine in density-ratio space

The Wu-Benkeser metalearner (`dr_sl = TRUE`) replaces step 3 above.

**Core idea.**  Let r_1, ..., r_K be the density-ratio predictions from K
base learners.  Rather than finding NNLS weights on the probability scale, WB
finds weights beta = (beta_1, ..., beta_K) on the simplex that minimise

  L(beta) = mean[ s_i * log( sum_k beta_k * r_k(x_i) ) ]

where s_i = 1 - 2 * Y_i (so s_i = -1 for the shifted arm and +1 for the
natural arm).  This is a cross-entropy loss expressed directly in
density-ratio space; it arises from the log-likelihood of an exponential
tilt model.

Weights are found by BFGS on the softmax reparameterisation alpha_k such that
beta_k = exp(alpha_k) / sum_j exp(alpha_j), which enforces the simplex
constraint automatically.  If BFGS fails to converge, equal weights are
used as a fallback with a warning.

**The key distinction: where selection happens.**  All metalearners --
including WB -- combine base learner predictions into an ensemble.  The
difference is the scale on which learners are evaluated and selected.

Standard metalearners (NNLS, log-likelihood, etc.) evaluate base learners
on their **probability predictions**: learners are scored by how well they
classify natural-course vs shifted-course observations, and the ensemble
weights are chosen to minimise a loss on that probability scale.  The
resulting ensemble probability is then converted to a density ratio via
p / (1 - p).  This works well when all base learners output probabilities in
(0, 1), but it means that the selection criterion is one step removed from
the quantity of interest (the density ratio itself).

The WB metalearner evaluates and selects base learners **directly on the
density-ratio scale**, using a log-DR loss that is natural to that space.
This has two practical consequences:

1. Learners are selected for their density-ratio estimation accuracy, not
   their probability classification accuracy -- which is the more relevant
   criterion for downstream estimators.
2. Base learners that **directly output density ratios** (rather than
   probabilities) can be included in the library and combined under the same
   loss.  Direct density-ratio estimators such as KLIEP (Kullback-Leibler
   Importance Estimation Procedure), RuLSIF (relative unconstrained
   Least-Squares Importance Fitting), and parametric exponential-tilt models
   fit on the log-ratio scale are natural candidates.  These methods estimate
   the ratio r(x) = dP~(A|x) / dP(A|x) directly without going through a
   classification step, and are often more efficient when the ratio is far
   from 1 or the covariate dimension is high.

---

## Clipping in the WB pathway

The standard clipping applied to all SL predictions elsewhere in the
pipeline -- via `scale_info$clip` with the user-supplied `bounds` parameter
-- does **not** apply to WB base learners, because those learners return
density ratios (not probabilities) and there is no probability-space
analogue to clip against.

Clipping in the WB pathway is instead handled by `dr_floor` inside
`method.WB_dr()`:

```{r dr-floor}
# dr_floor floors every base-learner density-ratio prediction before
# the log is taken.  Prevents log(0) during BFGS and in computePred.
method.WB_dr(dr_floor = 1e-10)
```

Pass this to `density_ratio()` via the `method_g` argument:

```{r wb-call}
wr <- density_ratio(
  ...,
  dr_sl    = TRUE,
  method_g = method.WB_dr(dr_floor = 1e-10)
)
```

---

## Custom DR-returning wrappers (required)

**Standard SuperLearner wrappers are incompatible with `dr_sl = TRUE`.**
Wrappers such as `SL.glm` or `SL.xgboost` return probabilities in (0, 1).
When WB's `computePred` multiplies them as density ratios, the result is
numerically meaningless.

The WB pathway requires base learners whose `predict` method returns a
**density ratio**, not a probability.  The interface a DR-returning wrapper
must satisfy:

- `SL.my_dr_learner(Y, X, newX, family, ...)` is called during SuperLearner
  cross-validation.  `Y` is the 0/1 arm indicator (0 = natural, 1 = shifted).
  `X` and `newX` are the covariate matrices.
- The returned `pred` field must contain **density-ratio predictions**
  r-hat(x) = dP~(A|x) / dP(A|x) evaluated at the rows of `newX`.
- The returned `predict` S3 method must also return density ratios (not
  probabilities).
- Predictions should be non-negative; apply `dr_floor` or equivalent
  flooring inside the wrapper before returning.

A concrete example structure:

```{r wrapper-example}
SL.my_kliep_learner <- function(Y, X, newX, family, obsWeights, ...) {
  # Fit a kernel density-ratio model (KLIEP, RuLSIF, etc.)
  # Y = 0 for natural-course rows, Y = 1 for shifted rows.
  fit <- my_kliep_fit(X[Y == 0, ], X[Y == 1, ])

  # Predictions must be density ratios, not probabilities
  dr_pred <- predict(fit, newX)
  dr_pred <- pmax(dr_pred, 1e-10)   # floor -- not the same as bounds clipping

  list(
    pred = dr_pred,
    fit  = structure(list(model = fit), class = "SL.my_kliep_learner")
  )
}

predict.SL.my_kliep_learner <- function(object, newdata, ...) {
  dr_pred <- predict(object$model, newdata)
  pmax(dr_pred, 1e-10)
}
```

Built-in DR-returning wrappers compatible with the WB pathway are planned
for version 1.0 of CausalState.

---

## When to prefer WB over NNLS

The WB metalearner may perform better than NNLS when:

- Base learners span very different density-ratio scales and NNLS
  averaging on the probability scale creates a poor ensemble.
- The density ratio is far from 1 (strong treatment shift), where
  probability-scale averaging introduces more distortion.
- You are including a direct density-ratio learner (e.g. KLIEP or RuLSIF)
  alongside classification learners, and want all learners combined on a
  common DR scale.

NNLS on the probability scale is generally more stable and is the
recommended default.  Use WB only when you have a specific reason and
appropriate DR-returning wrappers.

---

## Full example

```{r full-example}
library(CausalState)
library(SuperLearner)

# Assume sl_g_dr is a character vector of DR-returning wrapper names
# registered in the current R session.

wr_wb <- density_ratio(
  df              = patient_data,
  a_names         = "A",
  tmax            = 7L,
  baseline        = c("age", "sex"),
  tv_names        = c("L1", "L2"),
  sl_g            = sl_g_dr,           # DR-returning wrappers only
  dr_sl           = TRUE,
  method_g        = method.WB_dr(dr_floor = 1e-10),
  k               = 5L,
  inner_v         = 5L,
  v               = 5L,
  seed            = 1L,
  id              = "id",
  time            = "time",
  policy_spec_fun = policy_up
)

# Pass wr_wb to sdr() / itmle() as usual
res_sdr_wb <- sdr(
  df              = patient_data,
  weight_object   = wr_wb,
  ...
)
```

---

## References

Wu C, Benkeser D (2024). Nonparametric Efficient Estimation of Marginal
Structural Models using Targeted Machine Learning. arXiv:2408.10847.

Sugiyama M, Suzuki T, Kanamori T (2012). *Density Ratio Estimation in
Machine Learning*. Cambridge University Press.

Diaz I, Williams N, Hoffman KL, Schenck EJ (2021). Nonparametric Causal
Effects Based on Longitudinal Modified Treatment Policies. *JASA*
118(542):846-857.
